|

### 2021-05-28 04-Chp-CPRD-Analysis.Rmd
Submission of Chapter 4 to supervisors

# Primary analysis of lipid regulating agents and dementia {#cprd-analysis-heading}

@@

## Lay summary
Large-scale electronic health record (EHR) databases are created through the routine collection of information from patients. Under this system, whenever a patient attends their GP, their clinical data is recorded in a central database using a standard set of codes. These databases have substantial advantages for research studies, including the number of people they contain and the length of time for which participants are followed. This is particulary important when studying diseases such as dementia, which may begin to develop long before symptoms are seen.

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across England. Using this data, I examined the effect of treatment which lower cholesterol levels, such as statins, on the risk of dementia and related outcomes.

Lipid regulating agents were found to have no or minimal effect on Alzheimer's disease, but increased the risk of vascular dementia. This increase is very likely to be due to the presence of bias in the analysis despite best efforts to address it. However, this analysis still provides important sources of information which will be used in later chapters.

@@

This chapter details an analysis of a large population-based electronic health record dataset to investigate the relationship between lipid regulating agent use and dementia outcomes. The analysis aims to address two important limitations of the current evidence base as identified by the systematic review presented in Chapter \@ref(sis-rev-heading).

In the first instance, it explicitly examines vascular dementia as an outcome. This is particular important to the thesis as a whole, as the systematic review identified an absence of information on the effect of lipids on dementia risk.

In the second instance, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia. This approach was taken to provide a further distinct evidence source for the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

Previously know as the General Practice Research Database (GPRD), the Clinical Practice Research Datalink (CPRD) is a large population-based, electronic health record (EHR) database.[@herrett2015] The database has been collected primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the primary care records for more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

To avoid the ambiguity of interpreting free-text clinical notes and to allow for easy analysis of the resulting data, the CPRD collects data use a predefined coding system known as Read codes.[@booth1994] All clinical events can be identified by a specific code. The codes use a nested approach (see Table \@ref(tab:readExample-table)), with the initial characters defining broad diagnostic topics (e.g. Eu... - Mental and behavioural disorders), while subsequent digits provide addition information on the specific condition diagnosed (e.g. Eu001 - Dementia in Alzheimer's disease with late onset).

@@

<!--------------------------------------------------------------------------->
(ref:readExample-caption) Example of CPRD Read code hierarchy, showing how "Dementia in Alzheimer’s disease with late onset" is nested under the top-level of "Mental disorders". Broad topics are specified using the initial two alpha-numeric characters of the Read code, while subsequent characters are used to define specific conditions and context.

@@

Using this system, lists of relevant codes for each of the index events, exposures and outcomes used in the analysis were created.

@@

This analysis was a prospective cohort study using data from the CPRD. It included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cutoff were included in the original CPRD extract but were excluded from the analysis as records are expected to be more complete and reliable after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort the 1st January 1996.

@@

Participants were included in our study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a lipid-regulating agent (such as statins); a total cholesterol test result of >4 mmol/L; or an LDL-c test result of >2 mmol/L.

These index events allowed us to define a population of participants who were either at risk of hypercholesterolemia, as indicated by the elevated total or LDL cholesterol test results, or had already been diagnosed with it, as indicated by a diagnostic code/related prescription. This approach was employed in an attempt to reduce confounding by indication that we would expect to observe in the full cohort, because individuals not prescribed lipid-regulating agents likely be less healthy across a range of variables than those prescribed lipid-regulating agents, leading to a biased association been lipid-regulating agent use and dementia. Conditioning entry into the study into the study on being either “at-risk” or already diagnosed with hypercholesterolemia attempts to mitigate this bias.

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); or last registration date with their GP practice. Participants were removed from our sample if they were less than 40 years of age, had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with dementia before or on the date of the index event.

@@

We considered seven lipid-regulating drug classes based on groupings in the British National Formulary (BNF)[@wishart2017], namely: statins, fibrates, bile acid sequestrants, ezetimibe, nicotinic acid groups, ezetimibe and statin (representing one treatment containing both drugs, rather than the two classes being prescribed concurrently), and omega-3 fatty acid groups.

A participant’s drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class was stopped (defined as last prescription of the primary class being followed by at least six months of observation), added to (defined as a second drug class being prescribed before the last prescription of the initial class), or switched (defined as a second drug class being prescribed after the last prescription of the initial class).

@@

### Outcomes

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome (Supplementary Figure 1). When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (see Figure ). The diagnosis date of the outcome was determined by the first record of a relevant code.

Participants were followed up, and were censored at the earliest of date of diagnosis, death, date of transfer out of the study cohort or end of follow-up (29th February 2016).

In addition to our primary outcomes of interest, we created two additional control outcomes. The inclusion of control outcomes in observational analyses are a useful technique to assess the strength of uncontrolled confounding.[@lipsitch2010] Negative outcomes are those without a likely causal pathway between the exposure and outcome (see Figure \@ref()). Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, our confidence in the result for the outcome of interest is increased.

Muscular backpain was used as a negative control outcome in this analysis. Despite observational analyses suggesting a link between statins and muscle pain (as opposed to more serious complications such as myopathy), systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the effect of statin use on muscle pain[@herrett2021] have found no evidence of an effect.

Additionally, incident ischemic heart disease was included as a positive control outcome, given the well-established protective effect of lipid-lowering treatment on the risk of this condition.[@collins2016a]

@@

<!----------------------------------------------------------------------------->
(ref:negativeOutcome-cap) Causal diagram showing relationship between exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of an arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders (assuming $C$ has been adjusted for).

@@

@@

A range of additional variables were included in the analysis, intended to address differential distributions of potential confounding variables between those who had were prescribed an LRA and those who were not.

The analysis was adjusted for a range of baseline covariates. Demographic covariates included age and gender. Age was calculated at date of entry into the cohort. Socioeconomic status was proxied using the Index of Multiple Deprivation, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

Body mass index (calculated as $weight/height^2$), baseline total cholesterol and baseline LDL cholesterol measures were obtained, using the last recorded value prior to the index date. A variable indicating grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010) was included to allow for changes in prescribing trends across the lifetime of the cohort. To assess healthcare utilisation, I calculated the average annual number of consultations between the beginning of a patients data and their entry into the cohort

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new]

@@

Given that they contain an administrative data, collected for the purposes of patient management and care rather than academic research, missing data are a recognised issue in electronic health records databases.[@wells2013strategies]

<!-- Failing to adjust for missing data can. . .  -->

<!-- Complete case analyses are limited by . . . -->

<!-- Review materials from multiple imputation course -->

In this analysis, missing data were handled using a multiple imputation approach. Variables with missing observations were identified, and 20 imputed dataset.[@sterne2009a] Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006using] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) command in STATA16.

Missing data was only considered an issue for variables where a numerical test result was expected, (e.g. BMI) or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition, rather than being considered missing.[@wells2013strategies]

To investigate the impact of multiple imputation versus a complete case analysis,[@pigott2001review] where participants missing any covariate are dropped from the dataset, we preformed and compared the results of both approaches.[@hughes2019a]

To estimate the effect of statins on dementia outcomes, we used a Cox proportional hazards model, defined as:

<!-- TODO __Bit on why Cox models are slightly different from others, due to proportional hazards assumptions__ -->

<!-- Hazard ratios are -->

<!-- Test for non-proportionality are therefore very important, as models which fail -->

<!-- Problems with testing for non-proportional hazards when  -->

<!-- TODO Need to describe accounting for clustering here: id it somethings I did? See other thesis for clustering by practice -->

<!-- TODO Finish paragraph from fragments below -->

In order to explore whether the choice of time axis affects the results of the analysis,I ran two simple models, one using participants age as the time-scale and a second using absolute time-since-entry and adjusting for participant age.

<!-- TODO Need to actually run these models, but the important things is that they can be run on the unimputed data! -->

### Immortal time bias and time-varying treatment indicators

<!-- Potential biases included time varying confounding, selection bias due to censoring on death and  -->
<!-- We use a Cox proportional hazards model with a time-varying treatment indicator. -->

Immortal time bias covers two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to exposure is excluded, leading to the exposed and control groups being followed up from different time points.[@levesque2010a] In this case, events that occur prior to the exposure event are missing from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time prior to the exposure date, and any events occurring within it, is inappropriately assigned to the exposed group. This second presentation appears to be common in the existing literature, as many of the studies included in the review in Chapter \@ref(sys-rev-heading) were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(#risk-of-bias-subheading))

To address the potential for immortal time bias in our analysis, we employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

<!-- TODO Discuss problems introduced by censoring on death -->

<!--------------------------------------------------------------------------->
(ref:immortalTimeBias-cap) Diagram illustrating immortal time bias

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures/sys-rev/prisma_flow.png"))
```
<!--------------------------------------------------------------------------->

<!-- TODO Re-run this analysis -->

Sensitivity cohorts were also created, stratifying first by year of entry into the cohort and second by remiving participants who may have been pregnant (as statins are contraindicated in pregnancy).

Additional analyses stratified by outcome and drug class were also performed.

<!-- TODO Described sensitivity analyses reported below here -->

<!-- Make sure to include sex as a sensitivity analysis, given that that is the purpose of the IPD analysis -->

@@

Exploring the results when using a different codelist

@@

Of the `r comma(attrition[1,1])` participants included in our extract, `r comma(attrition[14,1])` met the inclusion criteria (Figure \@ref(fig:cprdFlowchart)), with a total follow-up of `r comma(total_followup)` patient years at risk.

@@

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) Attrition of CPRD participants as the eligibility criteria were applied

@@

The median age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementia). The distribution of baseline characteristics across the drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

\blandscape
<!--------------------------------------------------------------------------->
(ref:cprdCharacteristics-caption) Patient characteristics by drug class

@@

<!-- FIX Table 1 is missing type 2 diabetes -->

A substantial majority (`r percentage.statins`) of participants prescribed a lipid-regulating agent were prescribed a statin. We excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent analysis based on the extremely small number of participants in these groups (Table \@ref(tab:cprdCharacteristics-table)). The "Ezetimibe and statins" treatment group represent those prescribed a treatment containing both ezetimibe and statins, rather than those where the two treats were prescribed concurrently.

@@

<!-- TODO Described the average length of time between patients being diagnosed with  -->

@@

<!--------------------------------------------------------------------------->
(ref:cprdSSA-caption) Participants who stopped, switched or added treatements by initial treatment type

@@

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`, and baseline LDL cholesterol was missing for `r ldltext`

@@

All results described below are from the fully adjusted model. The results of the age-only model are presented in Supplementary Figure 3, and demonstrate that adjustment for additional covariates had a minimal effect.

<!--------------------------------------------------------------------------->
(ref:cprdPrimary-cap) __Results from primary analyses of CPRD data:__

@@

<!-- TODO Cross-check with the decision flowchart to ensure that  -->

@@

As shown in Figure \@ref(fig:cprdPrimary), the results of the analysis do not support an effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared with no treatment, with the exception of fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

__Choice of time axis__

__Complete case versus imputed data__

__Unadjusted versus full adjusted models__

>>>>

# Primary analysis of lipid regulating agents and dementia outcomes {#cprd-analysis-heading}

@@

<!-- TODO Check lipid-regulating vs lipid regulating -->

<!-- TODO Check all fig/table cross-references evaluate -->

## Lay summary

Electronic health record (EHR) databases are large collections of patients data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standard set of codes. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the length of time for which participants are followed. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across the UK. Using this data, the analysis presented in this chapter examined the effect of treatment which lower cholesterol levels, such as statins, on the risk of dementia and related outcomes.

No evidence for an effect of lipid regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of drugs called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia with lipid regulating agent use.

This increased risk is unexpected, and is very likely to be due to the presence of bias in the analysis. This bias, called "confounding by indication", is caused when those who are prescribed a statin are more at risk of vascular dementia for a range of reasons, and so it appears as if statins are harmful. However, despite this limitation, the analysis presented provides an important source of information which will be used in later chapters.

@@

In this Chapter, I present the results of an analysis of a large population-based electronic health record dataset to investigate the relationship between lipid regulating agent (LRA) use and dementia outcomes.

The analysis aims to address important two limitations of the current evidence base as identified by the systematic review presented in Chapter \@ref(sis-rev-heading).

Firstly, it explicitly examines vascular dementia as an outcome. The systematic review presented in the previous chapter identified an evidence gap around the effect of lipid regulating agents on the risk of vascular dementia. As triangulation exercises require as many diverse sources of evidence as possible, this analysis provides a source of information on this outcome.

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia (as identified by the systematic review). Specifically, this involved a concerted effort to address immortal time bias though use of a Cox Proportional time-varying treatment indicator. This approach provides a evidence source at risk of a distinct set of biases due to the alternative analytical strategy that will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

Previously known as the General Practice Research Database (GPRD), the Clinical Practice Research Datalink (CPRD) is a large population-based, electronic health record (EHR) database.[@herrett2015] The database has been collecting primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the primary care records for more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

To avoid the ambiguity of interpreting free-text clinical notes and to allow for easy analysis of the resulting data, the CPRD primarily collects data using a predefined coding system known as Read codes.[@booth1994] All clinical events, included clinical test results and diagnoses, can be identified by a specific Read code. The codes use a nested approach (see Table \@ref(tab:readExample-table)), with the initial characters defining broad diagnostic topics (e.g. Eu... - Mental and behavioural disorders), while subsequent characters provide additional information on the specific condition diagnosed (e.g. Eu001 - Dementia in Alzheimer's disease with late onset).

@@

<!--------------------------------------------------------------------------->
(ref:readExample-caption) Example of CPRD Read code hierarchy, showing how "Dementia in Alzheimer’s disease with late onset" (_Eu001_) is nested under the top-level of "Mental disorders" (_Eu..._). Broad topics are specified using the initial two alpha-numeric characters of the Read code, while subsequent characters are used to define specific conditions and context.

@@

Lists of relevant codes for each of the index events, exposures and outcomes used in the analysis were created. Each of these are described in more detail in the following sections.

@@

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cutoff were included in the original CPRD extract obtained for this analysis. However, these were excluded from the analysis as data quality and reliability is thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

Participants were included in the study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a lipid-regulating agent (such as statins); a total cholesterol test result of >4 mmol/L; or an LDL-c test result of >2 mmol/L. These index events allowed me to define a population of participants who were either at risk of hypercholesterolemia, as indicated by the elevated total or LDL cholesterol test results, or had already been diagnosed with it, as indicated by a diagnostic code/related prescription.

@@

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); or last registration date with their GP practice. Participants were removed from the sample if they were less than 40 years of age, had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

We considered seven lipid-regulating drug classes based on groupings in the British National Formulary (BNF),[@wishart2017] namely: statins, fibrates, bile acid sequestrants, ezetimibe, nicotinic acid groups, ezetimibe and statin (representing one treatment containing both drugs, rather than the two classes being prescribed concurrently), and omega-3 fatty acid groups.

A participant’s drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class altered according to one of three criteria:

- __stopped__: defined as the last prescription of the primary class being followed by at least six months of observation; 
- __added__: defined as a second drug class being prescribed before the last prescription of the initial class; and 
- __switched__: defined as a second drug class being prescribed after the last prescription of the initial class.

@@

### Outcomes {#cprd-outcomes}

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (see Figure \@ref(fig:decisionTreeFig)). The diagnosis date of the outcome was determined by the first record of a relevant code.

<!----------------------------------------------------------------------------->
(ref:decisionTreeFig-cap) Decision tree for assigning dementia subtypes, based on the presence of Read codes in the patient's record. Note that an outcome of "Probable" or "Possible" Alzheimer's disease (AD) requires the absence of any vascular outcome codes.

(ref:decisionTreeFig-scap) Decision tree for assigning dementia subtypes

```{r decisionTree, include = FALSE}
```

```{r decisionTreeFig, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:decisionTreeFig-cap)', out.width='100%', fig.scap='(ref:decisionTreeFig-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","decision_tree.png"))
```
<!----------------------------------------------------------------------------->

Participants were censored at the earliest of date of: diagnosis; death; transfer out of the study cohort; or end of follow-up (29th February 2016).

A range of additional variables were included in the analysis, intended to address the different distributions of potential confounding variables between those who were prescribed an lipid-regulating agent and those who were not.

Demographic covariates adjusted for included age and gender. Age was calculated at date of entry into the cohort. Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived.  Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

Body mass index (a summary measure calculated as $weight/height^2$), baseline total cholesterol and baseline LDL cholesterol measures were obtained, using the last recorded value prior to the index date. A variable indicating grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010) was included to allow for changes in prescribing trends across the lifetime of the cohort. To assess healthcare utilisation, I adjusted for the average annual number of consultations between the beginning of a patients data and their entry into the cohort.

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new] Inclusion of this index allowed me to attempt to adjust for the general health of patients included in the analysis.

Missing data are a recognised issue in electronic health records databases,[@wells2013strategies] given that they contain administrative data, collected primarily for the purposes of patient management and care rather than academic research.

In this analysis, missing data were handled using a multiple imputation approach. Variables with missing observations were identified, and 20 imputed datasets were created.[@sterne2009a] Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006using] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) command in STATA16.

Missing data was only considered an issue for variables where a numerical test result was expected (e.g. BMI), or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition, rather than being considered missing.[@wells2013strategies]

Using multiple imputation to handle missing data is an alternative to a "complete case" approach,[@pigott2001review] where participants missing any covariate are dropped from the dataset. As a sensitivity analysis, I preformed and compared the results of both methods,[@hughes2019a] to investigate the impact of multiple imputation on the results.

A Cox proportional hazards (PR) model was used to estimate the effect of statins on dementia outcomes. Cox PR models are defined, in general terms, as:

A Cox PR model was chosen for this analysis as it inherently accounts for the length of time participants spend in each exposure group. Using this approach, time-at-risk can be properly attributed, thus mitigating the impact of immortal time bias. This is discussed in detail in the following section.

<!-- IDEA Bit on why Cox models are slightly different from others, due to proportional hazards assumptions. Test for non-proportionality are therefore very important, as models which fail-->

<!-- IDEA Need to describe accounting for clustering here: id it somethings I did? See other thesis for clustering by practice -->

<!----------------------------------------------------------------------------->
&nbsp;

### Immortal time bias and time-varying treatment indicators {#cprd-immortal-time-bias}

Immortal time bias covers two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to the exposure is excluded leading to the exposed and control groups being followed up from different time points.[@levesque2010a] For example, if the unexposed group are followed from a cholesterol test result, while the exposed group is followed from date of LRA prescription, any events that occur in the exposed group prior to the exposure event will be inappropriately excluded from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time prior to the exposure date, and any events occurring within it, is inappropriately assigned to the exposed group. This second presentation appears to be common in the existing literature, as several of the studies included in the systematic review presented in Chapter \@ref(sys-rev-heading) were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(risk-of-bias-subheading)).

This analysis is primarily concerned with the second presentation of immortal time bias, as all participants were followed from a common index date (earliest of: date of raised cholesterol test results; hypercholesterolemeia diagnosis; or LRA prescription). To address the potential for this second form of immortal time bias in the analysis, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

Under this approach, all patients start in the unexposed group, & contribute time-at-risk until they are prescribed a lipid regulating agent and move into the exposed group. Note, patients for whom prescription of a lipid regulating agent was the index event only contribute time to the exposed group (i.e. they enter the cohort and move into the exposed group on the same day).

<!--------------------------------------------------------------------------->
(ref:immortalTimeBias-cap) Diagram illustrating the two presentations of immortal time bias, as a selection bias (Panel A) and a misclassification bias (Panal B).

```{r, immortalTime, include = FALSE}
```

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","immortal_time.png"))
```
<!--------------------------------------------------------------------------->

<!-- IDEA Finish paragraph from fragments below -->

<!-- IDEA In order to explore whether the choice of time axis affects the results of the analysis,I ran two simple models, one using participants age as the time-scale and a second using absolute time-since-entry and adjusting for participant age. -->

The primary analysis examined the effect of a lipid regulating agent on dementia risk, stratified by outcome and drug class. To assess the robustness of the results, a number of sensitivity analyses were performed. These are described in the following sections.

#### Control outcomes

In addition to the primary outcomes of interest (described in Section \@ref(cprd-outcomes)), I extracted data on two additional control outcomes. The inclusion of control outcomes in observational analyses are a useful technique to assess the strength of uncontrolled confounding,[@lipsitch2010] and these outcomes are usually class as either "negative" or "positive" outcomes. Negative outcomes are those without a likely causal path between the exposure and outcome (see Figure \@ref(fig:negativeOutcome) for a directed acyclic graph describing an ideal negative outcome).

Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, confidence in the result for the outcome of interest is increased.

Due to the wealth of data available on statins as a lipid regulating agent, control outcomes were chosen in reference to this drug class. Muscular backpain was chosen as a negative control outcome in this analysis. Despite observational analyses suggesting a link between statins and muscle pain (as opposed to more serious complications such as myopathy),[@selva-ocallaghan2018] systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the effect of statin use on muscle pain[@herrett2021] have found no evidence of an effect. As such, if no evidence of an effect of statins on backpain is found in the analysis, then confidence in the results for the dementia outcomes is increased.

Additionally, incident ischemic heart disease was included as a positive control outcome, given the well-established protective effect of lipid-lowering treatment, via statins, on the risk of this condition.[@collins2016a] Similar to the backpain outcome, if the analysis strategy can reproduce this known protective association, this will provide evidence that potential confounders have been sufficiently adjusted for.

@@

<!----------------------------------------------------------------------------->
(ref:negativeOutcome-cap) Causal diagram showing relationship between exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of any arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders $U$ (assuming $C$ has been adjusted for).

@@

#### Impact of additional covariates

@@

#### Sensitivity cohorts

Two sensitivity cohorts were also created. The first stratified by year of entry into the cohort in an attempt to assess for time period effects. The second removed participants who may have been pregnant  (coded as under 55 and female) to assess the robustness of the estimates, as statins are contraindicated in pregnancy, .[@karalis2016]

<!-- IDEA Make sure to include sex as a sensitivity analysis, given that that is the purpose of the IPD analysis -->

@@

#### Statin properties

As detailed in the introduction, the properties of statins may be important in their effect, based on the ability of lipophilic statins to cross the blood brain barrier (see Section \@ref(hold) in the Introductory Chapter).

To explore whether any observed effect for the statin analysis varied by statin property, a sensitivity analysis was performed, stratifying by statin lipophilicity.

@@

#### Impact of dementia code lists

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a previous study published in 2009 by Smeeth _et al_.[@smeeth2009]

This previous study used a propensity matching approach to estimate the effect of statins on a range of outcomes, and a protective effect of statin use on all-cause dementia (`r estimate(0.81, 0.69,0.96,type = "HR")`) and non-AD dementia (`r estimate(0.82,0.69, 0.97,type = "HR")`), but no evidence of an effect on AD (`r estimate(0.81,0.49, 1.35, type="HR")`).

The code lists used were obtained through correspondence with the authors of that study, and are available for inspection (see Section \@ref(cprd-data-avail)).

@@

Of the `r comma(attrition[1,1])` participants included in the extract, `r comma(attrition[14,1])` met the inclusion criteria (Figure \@ref(fig:cprdFlowchart)), with a total follow-up of `r comma(total_followup)` patient years at risk.

@@

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) Attrition of CPRD participants as the eligibility criteria were applied. Most attrition was due to the absence of an index event of interest.

@@

The median participant age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementia). The distribution of baseline characteristics across the seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

\blandscape
<!--------------------------------------------------------------------------->
(ref:cprdCharacteristics-caption) Patient characteristics by drug class. Summary statistics are presented as "% (N)" unless otherwise specified in the variable name.

@@

A substantial majority (`r percentage.statins`) of participants prescribed a lipid-regulating agent were prescribed a statin. I excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent analysis based on the extremely small number of participants in these groups (N=127 & N=168 respectively; see Table \@ref(tab:cprdCharacteristics-table)). Note that the "Ezetimibe and statins" treatment group represent those prescribed a single treatment containing both ezetimibe and statins, rather than those where the two treatments were prescribed concurrently.

@@

<!-- IDEA Described the average length of time between patients being prescribed a statin and being diagnosed with dementia -->

@@

<!--------------------------------------------------------------------------->
(ref:cprdSSA-caption) Participants who stopped, switched or added treatments by initial treatment type.

@@

The stopping, addition and switching of drug classes was common across all drug classes (Table \@ref(tab:cprdSSA-table)).

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`, and baseline LDL cholesterol was missing for `r ldltext`

@@

The results of the primary analysis using the fully adjusted Cox proportional hazards model with participant age as the time scale are presented for each drug/outcome combination in Figure \@ref(fig:cprdPrimary).

For each outcome, the overall "Any drug" estimate was driven by the statin subgroup, based on it's large size relative to the other drug classes.

<!--------------------------------------------------------------------------->
(ref:cprdPrimary-cap) Results from primary analyses of CPRD data using the fully adjusted model and participant age as the time scale.

@@

@@

No evidence was found for an effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared with no treatment, with the sole exception of fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

The results of the series of sensitivity analyses performed are described in the following sections.

#### Complete case versus imputed data
